Gelation as a condensation frustrated by hydrodynamics and mechanical tension

### 2015-09-16 gel_cond/trunk/paper.tex
fig:hydro Probability of staying elongated for a triplet as panel a Histogram of Rg as inset

In Figure~\ref{fig:cell_vs_cap}b and c we show the final structures for two gels prepared at the same state point, one by our protocol, the other by introducing in a capillary an existing gel and thus shear melting it. The later is coarser, highlighting that shaking or shear melting protocols~\cite{lu2008gelation,Teece2011,Bartlett2012} are not equivalent to a quench. However even in our case we expect differences compared to a quench simulated via molecular or Brownian dynamics, because of the presence of a solvent mediating hydrodynamic interactions~\cite{Furukawa2010}. Thus our protocol might be very close to application relevant processes where the sudden addition of a component (e.g. salt, acid, depletion agent) induces aggregation in an otherwise stable suspension (e.g. milk).

In Figure~\ref{fig:cell_vs_cap}d shows a computer reconstruction from experimental coordinates of a typical gelation experiment close to the cluster-gel boundary. Immediately before ions enter the cell ($t=0$), the suspension is in a Wigner crystal state where the particles are far apart due to long range repulsion. As soon as the charges are screened, the particles begin to aggregate and form clusters that progressively connect to each other while coarsening to finally percolate through the field of view at $t=\SI{35}{\minute}$.

@@

In Figure~\ref{fig:phasediag}a, we divide the phase diagram in three regions based on the final state obtained by our protocol: at low polymer concentration ($c_p<\SI{0.2}{\gram\per\litre}$) sample fully relax to a fluid state; at very low colloid volume fraction ($\phi<0.05$) and high polymer concentration the particles condense into long-lived well separated clusters as observed in~\cite{Lu2006}; in the rest of the explored phase space we observe a long-lived space spanning network. We rationalise these observations by superimposing the spinodal line obtained by generalised free volume theory~\cite{Fleer2008}.

XXX discuss differences between theory and experiment.XXX

With our technique we are not limited to observation of the final state. We found that the path leading to the final state is not universal and depends on the colloid volume fraction even within the gel region. In Figure~\ref{fig:phasediag}b, we represent this path in the $(l_\text{max}/L, \bar{N}_C)$ plane. Here $\bar{N}_C$ is the mean number of neighbours, or coordination number quantifying the compacity of the structure. $l_\text{max}$ is the spatial extent of the largest cluster that we normalize by the size of the field of view $L$ to obtain a measure of the distance to percolation of the system. $l_\text{max}/L\approx 1$ indicates percolation. At high densities, percolation takes place first, within the first few $\tau_B$ after charge screening and then coarsening proceed. At low densities, percolation never takes place and we observe the compaction of individual clusters. In between, we observe the process detailed in Figure~\ref{fig:cell_vs_cap}d: formation of low-compacity clusters that then slowly connect together to build the percolating network. This process can take hundreds of $\tau_B$ and is competing with cluster compaction, as indicated by the oblique trajectory in Figure~\ref{fig:phasediag}b.

The initial similarity between the trajectories of dilute gels and cluster phase is striking, as if their dynamics were ruled by the same phenomenon. To confront this hypothesis, we compute the time dependent static structure factor $S(q,t)$. For all gel and cluster samples we observe the appearance of a low $q$ peak in $S(q)$, see Supplementary Figure XXX, which is characteristic of spinodal decomposition in a system of a conserved order parameter. To track the position of this peak, we compute the characteristic wave number $\langle q \rangle$ and show its temporal evolution in Figure~\ref{fig:phasediag}c. The curves for all samples follow a master curve coherent with spinodal decomposition kinetics: at short times $\langle q \rangle(t)$ shows a plateau indicating that the low $q$ peak builds up at constant wave number whereas, whereas at intermediate times we observe coarsening with $\langle q \rangle \sim t^{-1/3}$. Finally at longer times each sample deviates from the master curve to form a plateau indicating arrest. The more dilute samples arrest sooner.

@@

\begin{figure}
	\includegraphics{figs/hydro.pdf}
	\caption{\textbf{Hydrodynamics} \textbf{a} Evolution of triplet radii of gyration in a non percolating sample ($\phi=4~\%$, $c_p=\SI{1}{\gram\per\litre}$). The characteristic time to reach the stable compact cluster is much longer than the Brownian time. \textbf{b} Evolution of the aspect ratios of clusters of 4 particles and more in the same sample (dashed lines) and in a percolating sample ($\phi=8~\%$, $c_p=\SI{1.5}{\gram\per\litre}$, continuous lines) \textbf{c} Bond angle distribution relative to existing bonds (gray), to a future bond (red) or to a future bond involving an isolated particle (blue) obtained in the percolating sample. Future bonds are shifted to smaller angles, whereas gas adsorption takes place from larger angles. Insets sketch both cases, with present bonds drawn thick and future bonds drawn dotted.}
	\label{fig:hydro}
\end{figure}

>>>>

In Figure~\ref{fig:cell_vs_cap}b and c we show confocal microscopy images of the final structures for two gels prepared at the same state point, one by our protocol, the other by introducing in a capillary an existing gel and thus shear melting it. The later is coarser, highlighting that shaking or shear melting protocols~\cite{lu2008gelation,Teece2011,Bartlett2012} are not equivalent to a quench. However even in our case we expect differences compared to a quench simulated via molecular or Brownian dynamics, because of the presence of a solvent mediating hydrodynamic interactions~\cite{Furukawa2010}. Thus our protocol might be very close to application relevant processes where the sudden addition of a component (e.g. salt, acid, depletion agent) induces aggregation in an otherwise stable suspension (e.g. milk).

In Figure~\ref{fig:cell_vs_cap}d shows a computer reconstruction from experimental coordinates of a typical gelation experiment at a relatively low volume fraction ($\phi=7.5\%$). Immediately before ions enter the cell ($t=0$), the suspension is in a Wigner crystal state where the particles are far apart due to long range repulsion. As soon as the charges are screened, the particles begin to aggregate and form clusters that progressively connect to each other while coarsening to finally percolate through the field of view at $t=\SI{35}{\minute}$.

@@

In Figure~\ref{fig:phasediag}a, we divide the phase diagram in three regions based on the final state obtained by our protocol: at low polymer concentration ($c_p<\SI{0.2}{\gram\per\litre}$) sample fully relax to a fluid state; at very low colloid volume fraction ($\phi<0.05$) and high polymer concentration the particles condense into long-lived well separated clusters as observed in~\cite{Lu2006}; in the rest of the explored phase space we observe a long-lived space spanning network. As observed by others~\cite{Shah2003,Bergenholtz2003,lu2008gelation}, the theoretical spinodal line obtained by generalised free volume theory~\cite{Fleer2008} is a poor predictor of the gel boundary, especially at low volume fraction.

With our technique we can observe the path leading to the final state. To characterise this path, we compute the instantaneous mean number of neighbours $\bar{N}_C$, or coordination number that quantifies the compacity of the structure. We also compute the the spatial extent of the largest cluster $l_\text{max}$ that we normalize by the size of the field of view $L$ to obtain a measure of the distance to percolation of the system. Figure~\ref{fig:phasediag}b shows system trajectory in the $(l_\text{max}/L, \bar{N}_C)$ plane for various colloidal volume fractions. At high densities, percolation takes place first, within the first few $\tau_B$ after charge screening and then coarsening proceed. At low densities, percolation never takes place and we observe the compaction of individual clusters. In between, we observe the process detailed in Figure~\ref{fig:cell_vs_cap}d: formation of low-compacity clusters that then slowly connect together to build the percolating network. This process can take hundreds of $\tau_B$ and is competing with cluster compaction, as indicated by the oblique trajectory in Figure~\ref{fig:phasediag}b. The path to gelation is thus not universal and depends on the colloid volume fraction even within the gel region.

The initial similarity between the trajectories of dilute gels and cluster phase is striking, as if their dynamics were ruled by the same phenomenon. To confront this hypothesis, we compute the time dependent static structure factor $S(q,t)$. For all gel and cluster samples we observe the appearance of a low $q$ peak in $S(q)$, see Supplementary Figure XXX, which is characteristic of spinodal decomposition in a system of a conserved order parameter. To track the position of this peak, we compute the characteristic wave number $\langle q \rangle$ and show its temporal evolution in Figure~\ref{fig:phasediag}c. The curves for all samples follow a master curve coherent with spinodal decomposition kinetics: at short times $\langle q \rangle(t)$ shows a plateau indicating that the low $q$ peak builds up at constant wave number, whereas at intermediate times we observe coarsening with $\langle q \rangle \sim t^{-1/3}$. Finally at longer times each sample deviates from the master curve to form a plateau indicating arrest. The more dilute samples arrest sooner.

@@

\begin{figure}
	\includegraphics{figs/hydro.pdf}
	\caption{\textbf{Hydrodynamics} \textbf{a} Probability of compaction of an elongated triplet in a non percolating sample ($\phi=4~\%$, $c_p=\SI{1}{\gram\per\litre}$). The continuous line is the best exponential fit of characteristic time $27\tau_B$. Inset: probability distribution of radii of gyration of triplets. \textbf{b} Evolution of the aspect ratios of clusters of 4 particles and more in the same sample (dashed lines) and in a percolating sample ($\phi=8~\%$, $c_p=\SI{1.5}{\gram\per\litre}$, continuous lines) \textbf{c} Bond angle distribution relative to existing bonds (gray), to a future bond (red) or to a future bond involving an isolated particle (blue) obtained in the percolating sample. Future bonds are shifted to smaller angles, whereas gas adsorption takes place from larger angles. Insets sketch both cases, with present bonds drawn thick and future bonds drawn dotted.}
	\label{fig:hydro}
\end{figure}
